---
title: "cellmig: quantifying cell migration with hierarchical Bayesian models"
author: "Simo Kitanovski (simo.kitanovski@uni-due.de)"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User Manual: cellmig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


# Background

```{r}
library(cellmig)
library(ggplot2)
library(ggforce)
library(patchwork)
library(skimr)
ggplot2::theme_set(new = theme_bw(base_size = 10))
```


# Data (raw)
This is how a typical cell migration data looks like $\rightarrow$ a table.

Each rows is a cell with the following features:

  * well = unique well ID (w1, w2, w3, etc.).
  * plate = unique plate ID (p1, p2, p3, etc.). A plate contains multiple wells.
  * compound = compound name (c1, c2, c3, etc.)
  * treatment dose (0, 1, 5, 10, low, mid, high, etc.)
  * v = Observed cell migration speed (numeric)
  * offset = binary (0 or 1). Indicates whether a treatment should be used 
  for batch correction across plates. By default offset = 0 (no correction). 
  Set to 1 for specific treatment groups (compound x dose) used as offsets. 
  Ensure that this treatment group appears on each plate.

```{r}
data("d", package = "cellmig")
d$offset <- 0
knitr::kable(d[1:10,], row.names = F, digits = 2)
```

These are the unique plate, well and group IDs.

```{r}
apply(X = d[,2:ncol(d)], MARGIN = 2, FUN = unique)
```

In this vignette we will use simulated data from:

  * four **plates**: p1, p2, p3 and p4
  * 90 **wells** per plate: w1, w2, ..., w360
  * 20 **cells** per well with their migration speeds `v`
  * the wells are treated with one of six **compounds** c1, c2, ..., c6 at 
  one of five **doses** d1, d2, ..., d5. Each pair of a compound and dose 
  is described as group: c1|d1, c2|d2, ... etc.


```{r, fig.width=8, fig.height=6}
ggplot(data = d)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v, group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```

# We can also show the **mean migration** per well (across all 20 cells each well)

```{r, fig.width=8, fig.height=6}
dm <- aggregate(v~well+plate+compound+dose+group, 
                data = d, FUN = mean)
ggplot(data = dm)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v, group = well), 
            size = 1.5, alpha = 0.7)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```


# Model fitting

```{r, fig.width=7, fig.height=3.5}
o <- cellmig(x = d,
             control = list(mcmc_warmup = 300,
                            mcmc_steps = 1000,
                            mcmc_chains = 4,
                            mcmc_cores = 4,
                            mcmc_algorithm = "NUTS",
                            adapt_delta = 0.8,
                            max_treedepth = 10))
```


# What are the overall effects of compounds and doses on cell migration?

```{r, fig.width=6, fig.height=3}
ggplot(data = o$posteriors$mu_group)+
  geom_line(aes(x = dose, y = mean, col = compound, group = compound))+
  geom_point(aes(x = dose, y = mean, col = compound))+
  geom_errorbar(aes(x = dose, y = mean, ymin = X2.5., ymax = X97.5., 
                    col = compound), width = 0.04)+
  ylab(label = "effect")
```



# Compare the dose-response `profiles` for different compounds

```{r, fig.width=9, fig.height=5}
get_profiles(x = o)+
  patchwork::plot_layout(widths = c(1, 1, 5))
```


# Compare the effects between treatment group

## Pairwise dot-plot comparison $\rightarrow$ x minus y axis

  * x/y-axis treatment groups (combinations of compounds and doses)
  * $\rho$: Difference between treatment groups at y-x axis.
  * $\pi$: probability of observing either a completely positive or negative $\rho$

```{r, fig.width=8, fig.height=6}
u <- get_pairs(x = o)
u$plot
```



## Violin plot based comparison

  * from_groups: vector of treatment groups to consider (combinations of 
  compounds and doses)
  * to_group: target treatment group
  * violins show the posterior distributions of the differences ($\rho$: 
  each element from `from_groups` vs. `to_group`).
  * label: probability, $\pi$, of observing completely positive or negative 
  $\rho$

```{r, fig.width=7, fig.height=2.5}
u <- get_violins(x = o, 
                 from_groups = c("c1|d1", "c1|d2", "c1|d3", "c1|d4", "c1|d5"), 
                 to_group = "c2|d1")
u$plot
```



# Posterior predictive checks (PPCs)

```{r, fig.width=7, fig.height=6}
g <- get_ppc_violins(x = o)
g
```


```{r, fig.width=6, fig.height=6}
g <- get_ppc_means(x = o)
g
```




# `migcell` can be used to simulate data

```{r}
sim_data <- sim(N_well_cells = 30,
                N_plate = 6,
                N_group = 10,
                N_well_reps = 5,
                kappa_mu = 2,
                kappa_sigma = 0.3,
                sigma_bio = 0.3,
                sigma_tech = 0.1,
                alpha_p = rep(0, 6),
                mu_group = rep(0, 10),
                offset = 4)

sim_data$compound <- sim_data$group
sim_data$dose <- "low"
```


We can show the newly simulated cell migration velocity (y-axis) for each cell 
(dot) at plate (color)

```{r, fig.width=7, fig.height=3}
ggplot(data = sim_data)+
  geom_sina(aes(x = paste0("group=", group), col = plate, y = v, 
                group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()
```


... or we can show the mean cell migration speed per well (dot) in each group 
(x-axis) and plate (color)

```{r, fig.width=7, fig.height=3}
ggplot(data = aggregate(v~well+group+plate, data = sim_data, FUN = mean))+
  geom_sina(aes(x = paste0("group=", group), col = plate, y = v, 
                group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()
```



# Model ref-fitting

```{r, fig.width=7, fig.height=3.5}
osd <- cellmig(x = sim_data,
               control = list(mcmc_warmup = 300,
                              mcmc_steps = 1000,
                              mcmc_chains = 4,
                              mcmc_cores = 4,
                              mcmc_algorithm = "NUTS",
                              adapt_delta = 0.8,
                              max_treedepth = 10))
```



```{r}
osd$posteriors$alpha_p
```

```{r}
osd$posteriors$mu_group
```


```{r}
# kappa_mu = 2, kappa_sigma = 0.5,
osd$posteriors$kappa_mu
osd$posteriors$kappa_sigma
```


```{r}
# sigma_bio = 0.3, sigma_tech = 0.1,
osd$posteriors$sigma_bio
osd$posteriors$sigma_tech
```





# `migcell` can be used to simulate data

```{r}
sim_data <- sim(N_well_cells = 30,
                N_plate = 6,
                N_group = 10,
                N_well_reps = 5,
                kappa_mu = 2,
                kappa_sigma = 0.3,
                sigma_bio = 0.3,
                sigma_tech = 0.1,
                alpha_p = rep(0, 6),
                mu_group = rep(0, 10),
                offset = 4)

sim_data$compound <- sim_data$group
sim_data$dose <- "low"
```


We can show the newly simulated cell migration velocity (y-axis) for each cell 
(dot) at plate (color)

```{r, fig.width=7, fig.height=3}
ggplot(data = sim_data)+
  geom_sina(aes(x = paste0("group=", group), col = plate, y = v, 
                group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()
```


... or we can show the mean cell migration speed per well (dot) in each group 
(x-axis) and plate (color)

```{r, fig.width=7, fig.height=3}
ggplot(data = aggregate(v~well+group+plate, data = sim_data, FUN = mean))+
  geom_sina(aes(x = paste0("group=", group), col = plate, y = v, 
                group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()
```



# Prior predictive check


```{r}
sim_data <- sim(N_well_cells = 50,
                N_plate = 6,
                N_group = 10,
                N_well_reps = 5,
                kappa_mu = 2,
                kappa_sigma = 0.3,
                sigma_bio = 0.3,
                sigma_tech = 0.1,
                alpha_p = rep(0, 6),
                mu_group = rep(0, 10),
                offset = 4)

sim_data$compound <- sim_data$group
sim_data$dose <- "low"
```



# Session Info

```{r}
sessionInfo()
```

