---
title: "cellmig: quantifying cell migration with hierarchical Bayesian models"
author: "Simo Kitanovski (simo.kitanovski@uni-due.de)"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User Manual: cellmig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


# Background

```{r}
library(cellmig)
library(ggplot2)
library(ggforce)
library(patchwork)
library(skimr)
ggplot2::theme_set(new = theme_bw(base_size = 10))
```


# Data (raw)

```{r}
data("d", package = "cellmig")
knitr::kable(d[1:10,], row.names = F, digits = 2)
```

These are the unique experiment, plate, well and group IDs.

```{r}
apply(X = d[,2:ncol(d)], MARGIN = 2, FUN = unique)
```

In this vignette we will use simulated data from:

  * two **experiments**: e1, e2
  * four **plates** (two per experiment): p1, p2, p3 and p4
  * 90 **wells** per plate: w1, w2, ..., w360
  * 20 **cells** per well with their migration speeds `v`
  * the wells are treated with one of six **compounds** c1, c2, ..., c6 at 
  one of five **doses** d1, d2, ..., d5. Each pair of a compound and dose 
  is described as group: c1|d1, c2|d2, ... etc.


```{r, fig.width=8, fig.height=6}
ggplot(data = d)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, shape = experiment,
                y = v, group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```

# We can also show the **mean migration** per well (across all 20 cells each well)

Notice the systematic differences in the mean migration speed between
the two experiments: e1 and e2.

```{r, fig.width=8, fig.height=6}
dm <- aggregate(v~well+plate+experiment+compound+dose+group, 
                data = d, FUN = mean)
ggplot(data = dm)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, shape = experiment,
                y = v, group = well), size = 1.5, alpha = 0.7)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```


# Migration speed (scaled)

```{r, fig.width=8, fig.height=6}
ggplot(data = d)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v/max(v), shape = experiment,
                group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed [scaled by maximum v]")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```


# Model fitting

```{r, fig.width=7, fig.height=3.5}
o <- cellmig(x = d,
             control = list(mcmc_warmup = 300,
                            mcmc_steps = 1300,
                            mcmc_chains = 4,
                            mcmc_cores = 4,
                            mcmc_algorithm = "NUTS",
                            adapt_delta = 0.8,
                            max_treedepth = 10))
```


# What are the overall effects of compounds and doses on cell migration?

```{r, fig.width=6, fig.height=3}
ggplot(data = o$s$mu_group)+
  geom_line(aes(x = dose, y = mean, col = compound, group = compound))+
  geom_point(aes(x = dose, y = mean, col = compound))+
  geom_errorbar(aes(x = dose, y = mean, ymin = X2.5., ymax = X97.5., 
                    col = compound), width = 0.04)+
  ylab(label = "effect")
```



# Compare the dose-response `profiles` for different compounds

```{r, fig.width=9, fig.height=5}
get_profiles(x = o)+
  patchwork::plot_layout(widths = c(1, 1, 5))
```


# Compare the effects between treatment group

## Pairwise dot-plot comparison $\rightarrow$ x minus y axis

  * x/y-axis treatment groups (combinations of compounds and doses)
  * $\delta$: Difference between treatment groups at y-x axis.
  * $\pi$: probability of observing either a completely positive or negative $\delta$

```{r, fig.width=8, fig.height=6}
u <- get_pairs(x = o)
u$plot
```



## Violin plot based comparison

  * from_groups: vector of treatment groups to consider (combinations of 
  compounds and doses)
  * to_group: target treatment group
  * violins show the posterior distributions of the differences ($\delta$: 
  each element from `from_groups` vs. `to_group`).
  * label: probability, $\pi$, of observing completely positive or negative 
  $\delta$

```{r, fig.width=7, fig.height=2.5}
u <- get_violins(x = o, 
                 from_groups = c("c1|d1", "c1|d2", "c1|d3", "c1|d4", "c1|d5"), 
                 to_group = "c2|d1")
u$plot
```



# Posterior predictive checks (PPCs)

```{r, fig.width=8, fig.height=5}
g <- get_ppc(x = o)
g
```


# `migcell` can be used to simulate data

```{r}
sim_data <- sim(N_well_cells = 20,
                N_experiment = 2,
                N_plate = 3,
                N_group = 6,
                N_well_reps = 3,
                shape = 4,
                sigma_bplate = 0.3,
                sigma_wplate = 0.1,
                alpha_plate = matrix(data = c(1, 1.2, 1, 1, 1, 1), 
                                     nrow = 3, ncol = 2),
                alpha_experiment = c(1, 1),
                mu_group = c(-1, -0.5, -0, 0.5, 1, 2))
```

```{r, fig.width=6, fig.height=3}
ggplot(data = sim_data)+
  geom_sina(aes(x = paste0("group=", group), col = plate, y = v, 
                group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))
```



# Session Info

```{r}
sessionInfo()
```

