---
title: "Simulating data with cellmig"
author: "Simo Kitanovski (simo.kitanovski@uni-due.de)"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User manual: data simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


```{r}
library(cellmig)
library(ggplot2)
library(ggforce)
library(patchwork)
library(skimr)
ggplot2::theme_set(new = theme_bw(base_size = 10))
```


# Motivation
In this vignette, we demonstrate the capabilities of cellmig to generate 
cell velocity data through two case studies:

1. Simulation from a `partially generative` model (user-specified parameters 
with some random components)
2. Simulation from a `fully generative` model (all parameters randomly 
generated)

# Data simulation from **partially generative** model
We generate data using an experimental design specified in the `control` list:

```{r}
control <- list(
  N_biorep = 4,     # Biological replicates (independent plates)
  N_techrep = 3,    # Technical replicates (wells per plate)
  N_cell = 30,      # Cells per well
  delta = c(0, -0.4, -0.2, -0.1, 0, 0.1, 0.2, 0.4),  # Treatment effects
  sigma_bio = 0.2,  # Biological variability
  sigma_tech = 0.05,# Technical variability
  offset = 1,       # Systematic offset (explain usage if relevant)
  prior_alpha_p_M = 1.7,    # Mean for plate effects
  prior_alpha_p_SD = 0.5,   # SD for plate effects
  prior_kappa_mu_M = 1.7,   # Mean for kappa location
  prior_kappa_mu_SD = 0.5,  # SD for kappa location
  prior_kappa_sigma_M = 0,  # Mean for kappa scale
  prior_kappa_sigma_SD = 0.3# SD for kappa scale
)
```

Experimental design parameters:

  * 4 biological replicates (independent experimental plates, $p = \{1,2,3,4\}$)
  * 3 technical replicates per biological replicate (wells per plate)
  * 30 cells per well
  * 8 treatment groups ($t = \{1,...,8\}$) with fixed effects 
  $\delta_t$ = [0, -0.4, -0.2, -0.1, 0, 0.1, 0.2, 0.4]
  * Biological replicate variability: $\sigma_\text{bio}}$ = 0.2
  * Technical replicate variability: $\sigma_\text{tech}}$ = 0.05

## Model specification

### Setting $\alpha_p$
I assume that there are strong plate-specific batch effects, which is what
we typically observe in real experiments. To generate such effects, we draw 
random values for the parameter $\alpha_p$ 

$\alpha_p \sim \text{Normal}(1.7, 0.5)$ 

where $\alpha_p$ represents the plate-specific offset. The mean (defined by 
parameter `prior_alpha_p_M`=1.7) and standard deviation (defined by parameter 
`prior_alpha_p_SD`=0.5) of the normal distribution are defined in `control`.

### Setting $\kappa_w$
The velocities in each well are gamma distributed, and the gamma distribution
is defined by a rate and shape parameter. The shape parameter for well $w$, 
$\kappa_w$, controls the form or skewness of the distribution and, to some 
extent, the variability in the data.

We simulate the values of $\log(\kappa_w)$ by drawing randomly from a 
normal distribution with mean ($\mu_{\kappa}$) and standard deviation 
(${\sigma}_{\kappa}$). The values of $\mu_{\kappa}$ and ${\sigma}_{\kappa}$ 
can either be fixed to point values, or be drawn from two normal distributions 
whose means and standard deviations are described by the inputs `prior_kappa_mu_M`, `prior_kappa_sigma_M`, `prior_kappa_mu_SD`, and `prior_kappa_sigma_SD`. To fix 
the values of $\mu_{\kappa}$ and ${\sigma}_{\kappa}$ to point estimates set the 
scales to small values (e.g., set in `control` `prior_kappa_mu_SD`=0.01 and 
`prior_kappa_sigma_SD`).

### Defining the offset
To do batch effect correction across plates we have to select one of treatment
as a reference. We do this by setting the parameter `offset` = $x$ in the
`control`, where $x \in t$. In this example `offset=1`.

## Generates dataframe with simulated velocities

```{r}
# set seed for reproducible results
set.seed(seed = 1252)
# simulate data
y_p <- gen_partial(control = control)  
```

```{r}
# see data structure
str(y_p$y)
```

We can visualize data data as a scatterplot with cells as dots

  * y-axis = velocity
  * x-axis = treatment group
  * color = plate identifier

```{r, fig.width=6, fig.height=3}
ggplot(data = y_p$y)+
  geom_sina(aes(x = paste0("t=", group), col = paste0("p=", plate), 
                y = v, group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_color_discrete(name = "plate")+
  scale_y_log10()
```

... or we can just show the plate-specific means (below). Notice, the 
systematic batch effects between plates.

```{r, fig.width=6, fig.height=3}
ggplot(data = aggregate(v~well+group+plate, data = y_p$y, FUN = mean))+
  geom_sina(aes(x = paste0("t=", group), col = paste0("p=", plate), 
                y = v, group = well), size = 1)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()+
  scale_color_discrete(name = "plate")
```

### Generated data can be **analyzed** with `cellmig`

```{r, fig.width=7, fig.height=3.5}
sim_data <- y_p$y

# format simulated data to be used as input in cellmig
sim_data$well <- as.character(sim_data$well)
sim_data$compound <- as.character(sim_data$compound)
sim_data$plate <- as.character(sim_data$plate)
sim_data$offset <- 0
sim_data$offset[sim_data$group=="1"] <- 1


osd <- cellmig(x = sim_data,
               control = list(mcmc_warmup = 300,
                              mcmc_steps = 1000,
                              mcmc_chains = 3,
                              mcmc_cores = 3,
                              mcmc_algorithm = "NUTS",
                              adapt_delta = 0.8,
                              max_treedepth = 10))
```


### Compare simulated vs inferred parameters

Let's compare inferred (black) vs originally simulated (red) $\alpha_p$

```{r, fig.width=4, fig.height=2}
ggplot(data = osd$posteriors$alpha_p)+
  geom_point(aes(y = plate_id, x = exp(mean)))+
  geom_errorbarh(aes(y = plate_id, x = exp(mean), xmin = exp(X2.5.), 
                     xmax = exp(X97.5.)), height = 0.1)+
  geom_point(data = data.frame(v = exp(y_p$par$alpha_p), plate = 1:4),
             aes(y = plate, x = v), col = "red")+
  scale_y_continuous(breaks = 1:4)+
  xlab(label = expression(alpha[p]*"'"))
```

... and also $\delta_t$

```{r, fig.width=4, fig.height=2.5}
ggplot(data = osd$posteriors$mu_group)+
  geom_point(aes(y = group_id, x = mean))+
  geom_errorbarh(aes(y = group_id, x = mean, xmin = X2.5., xmax = X97.5.), height = 0.1)+
  geom_point(data = data.frame(delta = c(-0.4, -0.2, -0.1, 0, 0.1, 0.2, 0.4), group_id = 1:7),
             aes(y = group_id, x = delta), col = "red")+
  xlab(label = expression(delta[t]))+
  ylab(label = "treatment")+
  scale_y_continuous(breaks = 1:8)
```

# Data simulation from **fully generative** model 
We can also use `cellmig` to generate cell velocities from the prior 
distributions of the parameters $\rightarrow$ data from prior predictive 
distribution

In this scenario, the parameters, such as $\delta_t$, will not be fixed to
point values but rather generated from their priors.

## Model specification

```{r}
control <- list(N_biorep = 4, 
                N_techrep = 3, 
                N_cell = 30, 
                N_group = 8,
                prior_alpha_p_M = 1.7,
                prior_alpha_p_SD = 1.0,
                prior_kappa_mu_M = 1.7,
                prior_kappa_mu_SD = 1.0,
                prior_kappa_sigma_M = 0,
                prior_kappa_sigma_SD = 1.0,
                prior_sigma_bio_M = 0.0,
                prior_sigma_bio_SD = 1.0,
                prior_sigma_tech_M = 0.0,
                prior_sigma_tech_SD = 1.0,
                prior_mu_group_M = 0.0,
                prior_mu_group_SD = 1.0)
```



```{r}
# generate data
y_f <- gen_full(control = control)
```

The generate data has a similar structure as before:

```{r}
str(y_f)
```



## Compare density of velocities from partially vs. fully generative models

```{r, fig.width=4, fig.height=4}
w <- data.frame(v_f = y_f$y$v, v_p = y_p$y$v)
 ggplot(data = w)+
    geom_point(aes(x = v_f, y = v_p), size = 0.5)+
    geom_density_2d(aes(x = v_f, y = v_p), col = "orange")+
    scale_x_log10(name = "Velocity from fully generative model", limits = c(0.01, 10^4))+
    scale_y_log10(name = "Velocity from partially generative model", limits = c(0.01, 10^4))+
    annotation_logticks(base = 10, sides = "lb")+
    theme_bw(base_size = 10)
```


# Session Info

```{r}
sessionInfo()
```

