---
title: "cellmig: quantifying cell migration with hierarchical Bayesian models"
author: "Simo Kitanovski (simo.kitanovski@uni-due.de)"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{User Manual: cellmig}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Background

```{r}
library(cellmig)
library(ggplot2)
library(ggforce)
library(patchwork)
library(skimr)
ggplot2::theme_set(new = theme_bw(base_size = 10))
```

# Background

Cell migration is essential for many vital biological functions, for
instance division or differentiation, and it impacts pathological
processes including cancer metastasis. High-throughput tracking of cells
with time-lapse microscopy followed by the acquisition of images at ﬁxed
time intervals facilitates the analysis of cell velocity across many wells 
treated under different biological conditions, followed by computational 
image processing to extract quantitative cell velocities and to detect 
changes following perturbations such as drug treatments. However, these 
workflows generate considerable technical noise (e.g., imaging artifacts, 
unprecise pipetting) and biological variability (e.g., cell-to-cell
heterogeneity), and therefore technical and biological replicates are
necessary, leading to large, hierarchically structured datasets: cells
are nested within technical replicates that are nested within biological
replicates.

Current statistical analyses of such data rely on qualitative approaches, 
or inadequate statistical methods that usually ignore the hierarchical 
structure of the data and fail to explicitly quantify uncertainty arising 
from technical or biological variability. To address this gap, we present 
cellmig, an R package implementing Bayesian hierarchical models for 
migration analysis. cellmig quantifies condition-specific velocity changes 
(e.g., drug effects) while modeling nested data structures and technical 
artifacts, providing uncertainty-aware estimates through credible intervals. 
It further enables synthetic data generation for experimental design
optimization. 

# Data

This is how a typical cell migration data looks like $\rightarrow$ a
table.

Each rows is a cell with the following features:

-   well = unique well ID (w1, w2, w3, etc.).
-   plate = unique plate ID (p1, p2, p3, etc.). Each plate is a biological
    replicate. A plate contains multiple wells, some of which are treated 
    with the same compound and dose (technical replicates)
-   compound = compound name (c1, c2, c3, etc.)
-   dose = compound concentration (0, 1, 5, 10, low, mid, high, etc.)
-   v = Observed cell migration velocity (numeric)
-   offset = binary (0 or 1). Indicates whether a treatment should be
    used for batch correction across plates. By default offset = 0 (no
    correction). Set to 1 for specific treatment groups (compound x
    dose) used as offsets. Ensure that this treatment group appears on
    each plate.

```{r}
data("d", package = "cellmig")
str(d)
```

In this vignette we will use simulated data from:

-   **plates** ($p$): 1, 2, ... , 3
-   **wells** ($w$): 1, 2, ... , 378
-   **cells** per well with their migration velocity `v`
-   wells are treated with **compounds** 1, 2, ..., 6 at **dose** 1, 2,
    ..., 7.
-   combination of a compound and dose is a **treatment** ($t$)
    $\rightarrow$ 1, 2, ..., 42.


Let's visualize the data. Each dot represents a cell with its velocity on the y-axis. Each facet corresponds to a compound. The x-axis represents the dose. 
There are three plates, indicated by color. Four technical replicates (wells), analyzed on the same plate, are stacked next to each other.

```{r, fig.width=7, fig.height=6}
ggplot(data = d)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v, group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  ylab(label = "migration velocity")+
  xlab(label = '')+
  scale_color_grey()+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "l")
```

## **Mean migration velocity** per well

```{r, fig.width=7, fig.height=6}
dm <- aggregate(v~well+plate+compound+dose, data = d, FUN = mean)
ggplot(data = dm)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v, group = well), 
            size = 1.5, alpha = 0.7)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  ylab(label = "migration velocity")+
  xlab(label = '')+
  scale_color_grey()+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "l")
```

# Model fitting
Next, we fit the Stan model employed by cellmig.

```{r, fig.width=7, fig.height=3.5}
o <- cellmig(x = d,
             control = list(mcmc_warmup = 300,
                            mcmc_steps = 1000,
                            mcmc_chains = 3,
                            mcmc_cores = 3,
                            mcmc_algorithm = "NUTS",
                            adapt_delta = 0.8,
                            max_treedepth = 10))
```

## What are the **overall treatment effects** (parameter $\delta$) on cell migration velocity?

* Dot: Posterior mean of $\delta$
* Error bar: 95% highest density interval (HDI) of $\delta$
* $\exp(\delta)$: Fold change in cell velocity relative to control

```{r, fig.width=6, fig.height=3.3}
ggplot(data = o$posteriors$mu_group)+
  geom_line(aes(x = dose, y = mean, col = compound, group = compound))+
  geom_point(aes(x = dose, y = mean, col = compound))+
  geom_errorbar(aes(x = dose, y = mean, ymin = X2.5., ymax = X97.5., 
                    col = compound), width = 0.1)+
  ylab(label = expression("Overall treatment effect ("*delta*")"))+
  theme(legend.position = "top")
```

## Compare the dose-response `profiles` for different compounds

```{r, fig.width=9, fig.height=5}
get_dose_response_profile(x = o)+
  patchwork::plot_layout(widths = c(.7, 1, 4))
```

## Compare the treatment group `profiles`

```{r, fig.width=4, fig.height=6}
get_treatment_profile(x = o,
                      hc_link = "average",
                      hc_dist = "euclidean",
                      select_gs = get_groups(o)$group,
                      B = 1000)
```

# Compare the effects between treatment group

## Pairwise dot-plot comparison $\rightarrow$ x minus y axis

-   x/y-axis treatment groups (combinations of compounds and doses)
-   $\rho$: Difference between treatment groups at y-x axis.
-   $\pi$: probability of observing either a completely positive or
    negative $\rho$

```{r, fig.width=12, fig.height=6}
u <- get_pairs(x = o)
u$plot
```

## Violin plot based comparison

-   from_groups: vector of treatment groups to consider (combinations of
    compounds and doses)
-   to_group: target treatment group
-   violins show the posterior distributions of the differences ($\rho$:
    each element from `from_groups` vs. `to_group`).
-   label: probability, $\pi$, of observing completely positive or
    negative $\rho$

```{r, fig.width=7, fig.height=2.5}
get_groups(x = o)
u <- get_violins(x = o, 
                 from_groups = get_groups(x = o)$group,
                 to_group = "C2|D1")
u$plot
```

# Posterior predictive checks (PPCs)

```{r, fig.width=5, fig.height=6}
g <- get_ppc_violins(x = o)
g+scale_y_log10()
```

```{r, fig.width=4, fig.height=4}
g <- get_ppc_means(x = o)
g
```

```{r, fig.height=2, fig.width=6}
g_alpha_p <- ggplot(data = o$posteriors$alpha_p)+
  geom_errorbarh(aes(y = plate_id, x = mean, xmin = X2.5., xmax = X97.5.),
                 height = 0.1)+
  geom_point(aes(y = plate_id, x = mean))

g_sigma <- ggplot()+
  geom_errorbarh(data = o$posteriors$sigma_bio,
                 aes(y = "sigma_bio",
                     x = mean, xmin = X2.5., xmax = X97.5.),
                 height = 0.1)+
  geom_errorbarh(data = o$posteriors$sigma_tech,
                 aes(y = "sigma_tech",
                     x = mean, xmin = X2.5., xmax = X97.5.),
                 height = 0.1)+
  geom_point(data = o$posteriors$sigma_bio,
             aes(y = "sigma_bio", x = mean))+
  geom_point(data = o$posteriors$sigma_tech,
             aes(y = "sigma_tech", x = mean))+
  ylab(label = '')

g_alpha_p|g_sigma
```

# Session Info

```{r}
sessionInfo()
```
