---
title: "Simulating data with cellmig"
author: "Simo Kitanovski (simo.kitanovski@uni-due.de)"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User manual: data simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


```{r}
library(cellmig)
library(ggplot2)
library(ggforce)
library(patchwork)
library(parallel)

ggplot2::theme_set(new = theme_bw(base_size = 10))
```


# Data simulation from **partially generative** model


```{r}
N_simulations <- 10
N_bioreps <- c(2, 4, 6, 8, 10)
N_techreps <- c(2, 4, 6, 8, 10)
N_cellreps <- c(10, 100, 1000)
deltas <- log(c(1, seq(from = 0.5, to = 2, by = 0.25)[-3]))
```


```{r}
run_configure <- function(b, N_bio, N_tech, N_cells) {
  control <- list(
    N_biorep = N_bio,
    N_techrep = N_tech,
    N_cell = N_cells,
    delta = deltas,   # Treatment effects
    sigma_bio = 0.2,  # Biological variability
    sigma_tech = 0.05,# Technical variability
    offset = 1,       # Systematic offset (explain usage if relevant)
    prior_alpha_p_M = 0,    # Mean for plate effects
    prior_alpha_p_SD = 0.3,   # SD for plate effects
    prior_kappa_mu_M = 1.7,   # Mean for kappa location
    prior_kappa_mu_SD = 0.5,  # SD for kappa location
    prior_kappa_sigma_M = 0,  # Mean for kappa scale
    prior_kappa_sigma_SD = 0.3# SD for kappa scale
  )
  
  seed_id <- sample(x = 1000+1:10^6, size = 1)
  set.seed(seed_id)
  
  y_p <- gen_partial(control = control)  
  
  sim_data <- y_p$y
  
  # format simulated data to be used as input in cellmig
  sim_data$well <- as.character(sim_data$well)
  sim_data$compound <- as.character(sim_data$compound)
  sim_data$plate <- as.character(sim_data$plate)
  sim_data$offset <- 0
  sim_data$offset[sim_data$group==control$offset] <- 1
  
  o <- cellmig(x = sim_data,
               control = list(mcmc_warmup = 300,
                              mcmc_steps = 1000,
                              mcmc_chains = 3,
                              mcmc_cores = 1,
                              mcmc_algorithm = "NUTS",
                              adapt_delta = 0.8,
                              max_treedepth = 10))
  
  og <- o$posteriors$mu_group
  og$real_delta <- deltas[-1]
  og$fp <- (og$real_delta >= og$X2.5. & og$real_delta <= og$X97.5.) & 
    !(og$X2.5. <= 0 & og$X97.5. >= 0)
  og$b <- b
  og$N_bio <- N_bio
  og$N_tech <- N_tech
  og$N_cells <- N_cells
  og$seed_id <- seed_id
  
  return(og)
}
```


```{r}
o <- mclapply(X = 1:10, 
              FUN = run_configure, 
              N_bio = 2, N_tech = 2, N_cells = 10,
              mc.cores = 6)

o <- do.call(rbind, o)
```

```{r}
ggplot(data = o)+
  facet_grid(N_bio~N_tech)+
  geom_point(aes(x = as.character(N_cells), y = exp(mean), col = fp, 
                 group = as.factor(exp(o$real_delta))),
             position = position_dodge(width = 1))
```



```{r}
# for(N_bio in N_bioreps) {
#   for(N_tech in N_techreps) {
#     for(N_cells in N_cellreps) {
#     }
#   }
# }
```




Experimental design parameters:

  * 4 biological replicates (independent experimental plates, $p = \{1,2,3,4\}$)
  * 3 technical replicates per biological replicate (wells per plate)
  * 30 cells per well
  * 8 treatment groups ($t = \{1,...,8\}$) with fixed effects 
  $\delta_t$ = [0, -0.4, -0.2, -0.1, 0, 0.1, 0.2, 0.4]
  * Biological replicate variability: $\sigma_\text{bio}}$ = 0.2
  * Technical replicate variability: $\sigma_\text{tech}}$ = 0.05

## Model specification

### Setting $\alpha_p$
I assume that there are strong plate-specific batch effects, which is what
we typically observe in real experiments. To generate such effects, we draw 
random values for the parameter $\alpha_p$ 

$\alpha_p \sim \text{Normal}(1.7, 0.5)$ 

where $\alpha_p$ represents the plate-specific offset. The mean (defined by 
parameter `prior_alpha_p_M`=1.7) and standard deviation (defined by parameter 
`prior_alpha_p_SD`=0.5) of the normal distribution are defined in `control`.

### Setting $\kappa_w$
The velocities in each well are gamma distributed, and the gamma distribution
is defined by a rate and shape parameter. The shape parameter for well $w$, 
$\kappa_w$, controls the form or skewness of the distribution and, to some 
extent, the variability in the data.

We simulate the values of $\log(\kappa_w)$ by drawing randomly from a 
normal distribution with mean ($\mu_{\kappa}$) and standard deviation 
(${\sigma}_{\kappa}$). The values of $\mu_{\kappa}$ and ${\sigma}_{\kappa}$ 
can either be fixed to point values, or be drawn from two normal distributions 
whose means and standard deviations are described by the inputs `prior_kappa_mu_M`, `prior_kappa_sigma_M`, `prior_kappa_mu_SD`, and `prior_kappa_sigma_SD`. To fix 
the values of $\mu_{\kappa}$ and ${\sigma}_{\kappa}$ to point estimates set the 
scales to small values (e.g., set in `control` `prior_kappa_mu_SD`=0.01 and 
`prior_kappa_sigma_SD`).

### Defining the offset
To do batch effect correction across plates we have to select one of treatment
as a reference. We do this by setting the parameter `offset` = $x$ in the
`control`, where $x \in t$. In this example `offset=1`.

## Generates dataframe with simulated velocities

```{r}
# set seed for reproducible results
set.seed(seed = 1252)
# simulate data
y_p <- gen_partial(control = control)  
```

```{r}
# see data structure
str(y_p$y)
```

We can visualize data data as a scatterplot with cells as dots

  * y-axis = velocity
  * x-axis = treatment group
  * color = plate identifier

```{r, fig.width=6, fig.height=3}
ggplot(data = y_p$y)+
  geom_sina(aes(x = paste0("t=", group), col = paste0("p=", plate), 
                y = v, group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_color_discrete(name = "plate")+
  scale_y_log10()
```

... or we can just show the plate-specific means (below). Notice, the 
systematic batch effects between plates.

```{r, fig.width=6, fig.height=3}
ggplot(data = aggregate(v~well+group+plate, data = y_p$y, FUN = mean))+
  geom_sina(aes(x = paste0("t=", group), col = paste0("p=", plate), 
                y = v, group = well), size = 1)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  scale_y_log10()+
  scale_color_discrete(name = "plate")
```

### Generated data can be **analyzed** with `cellmig`

```{r, fig.width=7, fig.height=3.5}
sim_data <- y_p$y

# format simulated data to be used as input in cellmig
sim_data$well <- as.character(sim_data$well)
sim_data$compound <- as.character(sim_data$compound)
sim_data$plate <- as.character(sim_data$plate)
sim_data$offset <- 0
sim_data$offset[sim_data$group=="1"] <- 1


osd <- cellmig(x = sim_data,
               control = list(mcmc_warmup = 300,
                              mcmc_steps = 1000,
                              mcmc_chains = 3,
                              mcmc_cores = 3,
                              mcmc_algorithm = "NUTS",
                              adapt_delta = 0.8,
                              max_treedepth = 10))
```


### Compare simulated vs inferred parameters

Let's compare inferred (black) vs originally simulated (red) $\alpha_p$

```{r, fig.width=4, fig.height=2}
ggplot(data = osd$posteriors$alpha_p)+
  geom_point(aes(y = plate_id, x = exp(mean)))+
  geom_errorbarh(aes(y = plate_id, x = exp(mean), xmin = exp(X2.5.), 
                     xmax = exp(X97.5.)), height = 0.1)+
  geom_point(data = data.frame(v = exp(y_p$par$alpha_p), plate = 1:4),
             aes(y = plate, x = v), col = "red")+
  scale_y_continuous(breaks = 1:4)+
  xlab(label = expression(alpha[p]*"'"))
```

... and also $\delta_t$

```{r, fig.width=4, fig.height=2.5}
ggplot(data = osd$posteriors$mu_group)+
  geom_point(aes(y = group_id, x = mean))+
  geom_errorbarh(aes(y = group_id, x = mean, xmin = X2.5., xmax = X97.5.), height = 0.1)+
  geom_point(data = data.frame(delta = c(-0.4, -0.2, -0.1, 0, 0.1, 0.2, 0.4), group_id = 1:7),
             aes(y = group_id, x = delta), col = "red")+
  xlab(label = expression(delta[t]))+
  ylab(label = "treatment")+
  scale_y_continuous(breaks = 1:8)
```

# Session Info

```{r}
sessionInfo()
```

