---
title: "cellmig: quantifying cell migration with hierarchical Bayesian models"
author: "Simo Kitanovski (simo.kitanovski@uni-due.de)"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User Manual: cellmig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


# Background

```{r}
library(cellmig)
library(ggplot2)
library(ggforce)
library(patchwork)
library(skimr)
ggplot2::theme_set(new = theme_bw(base_size = 10))
```


# Data (raw)
This is how a typical cell migration data looks like $\rightarrow$ a table.

Each rows is a cell with the following features:

  * well = unique well ID (w1, w2, w3, etc.).
  * plate = unique plate ID (p1, p2, p3, etc.). A plate contains multiple wells.
  * compound = compound name (c1, c2, c3, etc.)
  * treatment dose (0, 1, 5, 10, low, mid, high, etc.)
  * v = Observed cell migration speed (numeric)
  * offset = binary (0 or 1). Indicates whether a treatment should be used 
  for batch correction across plates. By default offset = 0 (no correction). 
  Set to 1 for specific treatment groups (compound x dose) used as offsets. 
  Ensure that this treatment group appears on each plate.

```{r}
data("d", package = "cellmig")
str(d)
```

In this vignette we will use simulated data from:

  * **plates** ($p$): 1, 2, ... , 4
  * **wells** ($w$): 1, 2, ... , 504
  * **cells** per well with their migration speed `v`
  * wells are treated with **compounds** 1, 2, ..., 6 at **dose** 1, 2, ..., 7. 
  * combination of a compound and dose is a **treatment** ($t$) 
  $\rightarrow$ 1, 2, ..., 42.

```{r, fig.width=8, fig.height=6}
ggplot(data = d)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v, group = well), size = 0.5)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```

# We can also show the **mean migration** per well (across all 20 cells each well)

```{r, fig.width=8, fig.height=6}
dm <- aggregate(v~well+plate+compound+dose, data = d, FUN = mean)
ggplot(data = dm)+
  facet_wrap(facets = ~paste0("compound=", compound), scales = "free_y", ncol = 2)+
  geom_sina(aes(x = as.factor(dose), col = plate, y = v, group = well), 
            size = 1.5, alpha = 0.7)+
  theme_bw()+
  theme(legend.position = "top",
        strip.text.x = element_text(margin = margin(0.03,0,0.03,0, "cm")))+
  xlab(label = "dose")+
  ylab(label = "migration speed")+
  scale_color_manual(values = c("steelblue", "orange", "forestgreen", "purple"))+
  guides(color = guide_legend(override.aes = list(size = 3)))+
  guides(shape = guide_legend(override.aes = list(size = 3)))
```


# Model fitting

```{r, fig.width=7, fig.height=3.5}
o <- cellmig(x = d,
             control = list(mcmc_warmup = 300,
                            mcmc_steps = 1000,
                            mcmc_chains = 3,
                            mcmc_cores = 3,
                            mcmc_algorithm = "NUTS",
                            adapt_delta = 0.8,
                            max_treedepth = 10))
```


# What are the overall effects of compounds and doses on cell migration?

```{r, fig.width=5, fig.height=3}
ggplot(data = o$posteriors$mu_group)+
  geom_line(aes(x = dose, y = mean, col = compound, group = compound))+
  geom_point(aes(x = dose, y = mean, col = compound))+
  geom_errorbar(aes(x = dose, y = mean, ymin = X2.5., ymax = X97.5., 
                    col = compound), width = 0.04)+
  ylab(label = "effect")
```



# Compare the dose-response `profiles` for different compounds

```{r, fig.width=9, fig.height=5}
get_dose_response_profile(x = o)+
  patchwork::plot_layout(widths = c(1, 1, 4))
```


# Compare the treatment group `profiles`

```{r, fig.width=4, fig.height=6}
get_treatment_profile(x = o,
                      hc_link = "average",
                      hc_dist = "euclidean",
                      select_gs = get_groups(o)$group,
                      B = 1000)
```


# Compare the effects between treatment group

## Pairwise dot-plot comparison $\rightarrow$ x minus y axis

  * x/y-axis treatment groups (combinations of compounds and doses)
  * $\rho$: Difference between treatment groups at y-x axis.
  * $\pi$: probability of observing either a completely positive or negative $\rho$

```{r, fig.width=14, fig.height=6}
u <- get_pairs(x = o)
u$plot
```



## Violin plot based comparison

  * from_groups: vector of treatment groups to consider (combinations of 
  compounds and doses)
  * to_group: target treatment group
  * violins show the posterior distributions of the differences ($\rho$: 
  each element from `from_groups` vs. `to_group`).
  * label: probability, $\pi$, of observing completely positive or negative 
  $\rho$

```{r, fig.width=7, fig.height=2.5}
get_groups(x = o)
u <- get_violins(x = o, 
                 from_groups = get_groups(x = o)$group,
                 to_group = "C2|D1")
u$plot
```



# Posterior predictive checks (PPCs)

```{r, fig.width=6, fig.height=6}
g <- get_ppc_violins(x = o)
g+scale_y_log10()
```


```{r, fig.width=4, fig.height=4}
g <- get_ppc_means(x = o)
g
```



```{r, fig.height=2, fig.width=6}
g_alpha_p <- ggplot(data = o$posteriors$alpha_p)+
  geom_errorbarh(aes(y = plate_id, x = mean, xmin = X2.5., xmax = X97.5.),
                 height = 0.1)+
  geom_point(aes(y = plate_id, x = mean))

g_sigma <- ggplot()+
  geom_errorbarh(data = o$posteriors$sigma_bio,
                 aes(y = "sigma_bio",
                     x = mean, xmin = X2.5., xmax = X97.5.),
                 height = 0.1)+
  geom_errorbarh(data = o$posteriors$sigma_tech,
                 aes(y = "sigma_tech",
                     x = mean, xmin = X2.5., xmax = X97.5.),
                 height = 0.1)+
  geom_point(data = o$posteriors$sigma_bio,
             aes(y = "sigma_bio", x = mean))+
  geom_point(data = o$posteriors$sigma_tech,
             aes(y = "sigma_tech", x = mean))+
  ylab(label = '')

g_alpha_p|g_sigma
```



# Session Info

```{r}
sessionInfo()
```

